---
title: "Intro to strings workshop - exercise solutions"
author: "Lindsay Katz"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message=FALSE}
library(tidyverse)
```

```{r}
# define fruits character vector to use for exercises
fruits <- c("cherries", "green grapes", "granny smith apples", "watermelon", "pear")
```

# stringr introductory exercises

Exercise: Using `str_detect`, detect which strings in the `fruits` vector contain the pattern "es".

```{r}
str_detect(fruits, "es")
```

Exercise: Using `str_count`, get a count for the number of letter "r"'s or "g"'s detected in each element of the `fruits` vector.

```{r}
str_count(fruits, "r|g")
```

Exercise: Using `str_extract_all`, extract all occurrences of the pattern "gr" from the `fruits` input vector.

```{r}
str_extract_all(fruits, "gr", simplify = T)
```

Exercise: Using `str_sub`, extract the first 4 characters of each string from `fruits` .

```{r}
str_sub(fruits, end=4)
```

Exercise: Using `str_replace`, replace the first vowel detected in each string from `fruits` with a question mark.

```{r}
str_replace(fruits, "a|e|i|o|u", "?")
```

Exercise: Using `str_to_upper`, convert each string in `fruits` to upper case.

```{r}
str_to_upper(fruits)
```

Exercise: Using `str_remove_all`, remove all spaces between words present in the `fruits` vector.

```{r}
str_remove_all(fruits, " ")
```

Exercise: Using `str_squish` and `str_to_title`, remove excess whitespace from the following string: "   Department       OF statiSTical   sciences ", and convert it to title case.

```{r}
str_squish(str_to_title("   Department       OF statiSTical   sciences "))
```

Exercise: Using `str_split`, split all strings in `fruits` into sub-strings wherever a space exists.

```{r}
str_split(fruits, " ", simplify = T)
```

# Using regex patterns with `stringr` exercises

Exercise: Using `str_detect` with an anchor and positive lookahead, detect the strings in `fruits` which start with the pattern "gr" which is followed by a letter "a".

```{r}
str_detect(fruits, "^gr(?=a)")
```

Exercise: Using `str_extract`, extract strings from `fruits` which are composed _only_ of between 4 and 8 letters. Hint: use anchors.

```{r}
str_extract(fruits, "^[:alpha:]{4,8}$")
```

# Parliamentary Debates example exercises

```{r}
# read in hansard data
hansard <- read_csv("../data/hansard_ex.csv", show_col_types = F)
```

Exercise: Using `str_count`, create a column of the count of digits found in each `name.id`.

```{r}
hansard %>% mutate(n_digits = str_count(name.id, "\\d")) %>% 
  select(name.id, n_digits)
```
  
Exercise: Create new variables containing the first and last initial for each MP. (Hint: use a positive lookbehind to extract the first initial)

```{r}
hansard %>% mutate(first_initial = str_extract(name, "(?<=, )[:upper:]"),
                   last_initial = str_sub(name, end=1)) %>% 
  select(name, first_initial, last_initial)
```

Before moving on, use the code from the workshop to split statements for the last speech in the `hansard` dataframe. We will complete the remaining exercises with this.

```{r}
# grab last speech
last_speech <- hansard %>% slice(n())

# extract unique strings to split body on, store as character vector
to_separate <- str_extract_all(last_speech$body, "[:upper:][:lower:]{1,2} [:alpha:]+:|[:upper:][:lower:]{1,2} [:alpha:]+ interjecting—") %>% 
  unlist() %>% unique()

# separate rows on character vector
last_speech <- separate_rows(last_speech, body, sep=paste0("(?=", to_separate, ")", collapse="|"))
```

Exercise: Add a flag variable called `interject` that takes on a value 1 if the body of text ends in the pattern "interjecting—", and 0 otherwise.

```{r}
last_speech <- last_speech %>% 
  mutate(interject = ifelse(str_detect(body, "interjecting—$"), 1, 0))
```

Exercise: For those rows where `interject==0`, remove the name pattern preceding the body of text (i.e. everything from the beginning of the string up to and including the colon).

```{r}
last_speech %>% mutate(body = ifelse(interject==0, str_remove(body, "^.+:"), body)) %>% 
  select(body)
```
